---
title: "R Data Wrangling and Manipulation: Part 2"
theme: readable
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 12
    fig_height: 7
---

```{r chunksetup, include=FALSE} 
library(fs)
library(here)
library(tidyverse)
library(tibble)
```

## The big picture 

Part 2 focuses on extending the base skills learned in Part 1. We are still interested in turning messy data into tidy data. This section will cover several common problems with messy data sets [Wickham 2014](https://vita.had.co.nz/papers/tidy-data.html). These are 

1. Column headers that are values, not variable names. 
2. Overloading variables into a single column. 
3. Storing variables in both rows and columns. 
4. Storing multiple types of units in the same data frame. 
5. Storing a single type in multiple tables. 

Within our discussion of (4) and (5), we will cover two table verbs in the tidyverse. In real data analysis projects, we will rarely have one data frame to do all of our analysis. 

## Column headers that are values, not variable names 

This messy data type is common in survey data, especially if you're working off of someone else's crosstabs. 

Imagine you're researching the housing crisis in California. You come across a dataset from Berkeley's [Institute for Governmental Studies](https://igs.berkeley.edu/research/berkeley-igs-poll) that has information on preferences for moving broken down by age group. 

Read in the dataset called `cal_housing.csv` in the data folder. 

```{r, message = F, warning = F}
cal_housing <- read_csv(here("data/cal_housing.csv"))
glimpse(cal_housing)
```

Our end goal is to make a visualization showing consideration of moving by age group. `ggplot2` presumes that we have tidy data to work. We need to *pivot* the non-variable columns into a two-column key-value pair to make our graph. This pivot operation will turn the age columns (e.g. `18-29`) into an age column where the age 18-29 is one of the possible values.

Recall from Part 1 that the appropriate tidyverse function for this operation is `pivot_longer()`. 

```{r}
cal_housing_long <- cal_housing %>% 
  pivot_longer(-leaving, 
               names_to = "age", 
               values_to = "percent_agreement")
head(cal_housing_long)
```

From here, we could make a bar graph showing response by age groups. 
```{r}
cal_housing_long %>% 
  ggplot(aes(x = leaving, y = percent_agreement, fill=age))+
  geom_col(position = 'dodge')+
  xlab("Moving Consideration Level")+
  ylab("Percent Agreement")+
  # colorblind friendly palette
  scale_fill_discrete(type = "viridis")+
  ggtitle("Californians Consideration to Leaving the State")

```

## Challenge 9 

The IGS poll also contains a question broken down by region (`cal_region.csv`). Build a pipeline that reads and tidies the data. For extra practice, make a similar graph to display the results. 

```{r}
## Fill in your solution here 
```

## Overloading variables into a single column 

Sometimes we have a dataset where our key column is a combination of multiple variables. This is a relatively standard raw data format in research projects that collect data at multiple sub-level sites over time. 

To illustrate the problem, consider the following dataset inspired by bird watchers. 

```{r}
birds <- read_csv(here("data/birds.csv"))
glimpse(birds)
```

We see that we have columns that combine two different variables, the species of bird ("eagle" and "warbler") and the location of a sighting ("nest" and "air"). When encountering this situation, we use a two-step process. 

1. Pivot to gather the non-variable columns 
2. Separate the overloaded column. 

In a workflow, it looks like this. 

```{r}
goodBirds <- birds %>% 
  pivot_longer(!c(county,year), 
               names_to = "sight",
               values_to = "n",
               values_drop_na = TRUE) %>%
  separate(sight, c("type", "location"), sep = "(?=[A-Z])")
glimpse(goodBirds)
```

The workflow above used a [regular expression](https://stringr.tidyverse.org/articles/regular-expressions.html) to split and keep the capital letter. Regular expressions are, in general, beyond the scope of this course. 

Another way to separate is by position. That call looks like this. 

```{r, eval = F}
separate(col = VAR, into = c("VAR1", "VAR2"), sep = NUMBER)
## example that splits on the first character 
separate(col = col, into = c("var1", "var2"), sep = 1)
```

Separating by position on datasets with a standard format can be helpful. Let's look at what happens when we separate by the 1st position. 

```{r}
sepEx <- birds %>% 
  pivot_longer(!c(county,year), 
               names_to = "sight",
               values_to = "n",
               values_drop_na = TRUE)%>%
  separate(sight, c("type", "location"),sep = 1)
glimpse(sepEx)
```

We can fix our incorrect columns with `mutate()`. The following is a brute force method to do this, but as you continue in your data wrangling journey, I strongly encourage you to seek out more efficient ways of solving this problem. 

```{r}
sepExFixed <- sepEx %>% 
  mutate(type = if_else(type == "e", "Eagle", "Warbler"),
         location = case_when(
           location == "agleNest"~"Nest",
           location == "agleAir"~"Air",
           location == "arblerNest"~"Nest",
           location == "arblerAir"~"Air"
         ))
head(sepExFixed)
```


## Challenge 10 

Edit the workflow presented above to split on the fifth character. Use `mutate` to replace the incorrect values with corrected and capitalized names. You can follow the brute force way shown above, or potentially investigate [str_replace_all()](https://stringr.tidyverse.org/reference/str_replace.html), [str_detect()](https://stringr.tidyverse.org/reference/str_detect.html), and [str_to_title()](https://stringr.tidyverse.org/reference/case.html) for options with string matching. 

```{r}


```


## Storing variables in both rows and columns 

Far less often, though far more frustratingly, variables can be stored in both rows and columns. Such datasets are surprisingly frequent in climate, geological, and health research. As a representative example of this problem, consider the hypothetical surveillance testing schedule for an athlete at Berkeley during the season. The data is called `testing.csv`

```{r}
testing <- read_csv(here("data/testing.csv"))
head(testing)
```

Note that we have variables in individual columns and a variable spread across multiple columns (the day variable), and variables spread across rows (the test variable is composed of two separate tests). 

In such a situation, we make use of both `pivot_longer()` and `pivot_wider()`. Our design pattern is to gather any variables spread across columns and then widen any columns with multiple variables within. 

```{r}
## Step 1 
step_1 <- testing %>% 
  pivot_longer(
    # an alternative method for selecting columns
    d1:d7, 
    names_to = "day",
    values_to = "result",
    values_drop_na = TRUE
  )%>%
  # remove the d from day and trim any white space
  # VERY USEFUL to avoid merge problems
  mutate(day = as.numeric(str_replace_all(day,"d", ""))%>%trimws())
head(step_1)
```

We now have a dataset close to a tidy dataset, except we have a problem. Two tests are being performed each day, and we only have one variable—time for step 2. 

```{r}
step_2 <- step_1 %>% 
  pivot_wider(
    names_from = test, 
    values_from = result
  )
glimpse(step_2)
```

Now we have one variable in each column, and each row represents a single day of testing. 

A trick that we can sometimes use occurs if we have data that looks like the following: 

```{r}
underscore <- read_csv(here("data/underscore.csv"))
glimpse(underscore)
```

This is wide data that we would like to transform to long data, but if we do a usual `pivot_longer()`, we will end up with the problem we want to avoid in this section. 

```{r}
# Bad
underscore %>% 
  pivot_longer(cols = -id,
               names_to = "observation",
               values_to = "value")
```

Here the dates and the results are merged into one value. However, in situations with a standard naming structure, we can use some of the built-in arguments of `pivot_longer()`.

```{r}
underscore %>% 
  pivot_longer(cols = -id,
               # special term indicates that column is split based
               # on character in name
               names_to = c("observation", ".value"),
               names_sep = "_",
               values_to = "value")

```

Once again, we get a tidy data frame that we can work with for future analyses. 


## Storing multiple types of units in the same data frame 

A fundamental principle of tidy data, and data wrangling in general, is to keep like for like in the same table. Datasets often violate this guideline when they have values collected at multiple levels on different types of observations. 

### Two Table Verbs

First, let's briefly review the two table verbs available in `dplyr`. The package breaks them up into three families. 

1. Mutating joins add new variables to one table from matching rows in another. 

```{r}
df1 <- tibble(a = c(2,4), b = c(4,6))
df2 <- tibble(a = c(3,2), x = 10, y = "b")
df3 <- tibble(other = c(2,2), yet_other = c(4,6), 
              x = 10, y = "b")
## inner_join() includes only observations that match both df1 and df2
df1 %>% 
  inner_join(df2)

## left_join() includes all observations in df1 regardless if they match or not. By far the most common join 
df1 %>% 
  left_join(df2)

## We can also specify the column(s) we want to specify common variables to join. 

## this will lead to an error 
df1 %>% 
  left_join(df3)

## These will return values 
df1 %>% 
  left_join(df3, by = c("a"= "other"))

df1 %>% 
  left_join(df3, by = c("a" = "other", 
                        "b"="yet_other"))

## right_join() includes all observations in df2. Equivalent to left_join() but with different ordered columns 
df1 %>% 
  right_join(df2)

## full_join() includes all observations from both df1 and df2
df1 %>% 
  full_join(df2)
```

Application: Making a balanced panel dataset 

A common type of dataset for analysis in many social science fields is [panel data](https://en.wikipedia.org/wiki/Panel_data). In a balanced panel, all units are observed each time unit. For example, if there were N units and T periods, the total number of rows would be NxT. 

The dataset for this exercise is `obs.csv`

```{r}
obs <- read_csv(here("data/obs.csv"))
```

We now take advantage of a useful function from `tidyr` called `expand()` along with functions that we have seen before. `expand()` generates all combinations of variables found in a dataset. We can pair it with `nesting()`, which only finds combinations already present in the data, or `crossing()`, which de-duplicates and sorts inputs. 

```{r}
panel <- obs %>%
  # the nesting function finds only the combinations that 
  # occur in the data
  expand(year,nesting(id))%>%
  left_join(obs)%>%
  # move the id column to the front of the data frame 
  # visually nice
  relocate(id)%>%
  arrange(id)
glimpse(panel)
```

2. Filtering joins match observations and are most useful for diagnosing mismatches in a join. 

```{r}
## semi_join()
df1 %>% 
  semi_join(df2)

## anti_join()
df1 %>% 
  semi_join(df2)
```

3. Set operations treat observations like sets and expect datasets to have the same variables. We are not covering these in this workshop. For more information, consult the `dplyr` manual. 

## Challenge 11: Bulk Recoding Values with a lookup table

Suppose you have a survey that asked three questions (var1,var2, var3) to fifteen subjects.

```{r}
survey_raw_data <-tibble(
  ID = LETTERS[1:15],
  var1 = sample(1:15, replace = F),
  var2 = sample(16:30, replace = F),
  var3 = sample(31:45, replace = F)
)
glimpse(survey_raw_data)
```

Survey data is likely to contain errors. Perhaps the enumerators filled in a question incorrectly. Maybe you have a particular value for missing data. As a subject matter expert, you realize that several fixes need to be made globally. You put them into a data frame. 

```{r}
fixes <- tibble(
  ID = c("A", "D", "F", "G"),
  original_var = c("var1", "var2", "var2", "var3"),
  original_response = as.integer(c(1, 28, 19, 34)),
  correct_response = as.integer(c(1000, 2800, 1900, 3400))
)
glimpse(fixes)
```

Here we have a problem of updating several variables based on a set of potentially different conditions. Each variable needs to be updated based on the ID, but we do not need to update every variable/ID combination. 

Build a workflow that converts the incorrect answers to correct answers. A solution to this problem makes use of `left_join()`, `mutate()`, `select()`, and both `pivot_longer()` and `pivot_wider()`

```{r}


```

## Binding Together Data Frames 

If you know that you have data frames that share columns or rows, you can efficiently turn them into a single data frame with `bind_rows()` or `bind_cols()`. The function that you use depends on the structure of your problem. 

```{r}
## An Example 
obs1 <- obs %>% 
  filter(id == "A")

obs2 <- obs %>% 
  filter(id == "B")

bind_rows(list(obs1, obs2))
```

## Storing a single type in multiple tables

It is common to find data values about a single observational unit spread out over multiple tables or files. Files are often split by another variable. For example, a set of medical records may be divided by patients. Voting records may be separated by counties or states. Economic data may be separated by economic sector. From a wrangling perspective, as long as there is a consistent format to these records, we can follow a three-step pattern: 

1. Read the files into a list of data frames
2. For each data frame, add a new column that records the original filename
3. Combine all the data frames into a single data frame. 

This pattern is so common that it is now built into the `readr` package. 

```{r, eval = F}
# Get appropriate files. 
# Here we assume that we have .csv files, but any 
# support file type will work
files <- fs::dir_ls(path = "/PATH", glob = "*pattern*csv")

## for example, maybe we have voting records by county 
files <- fs::dir_ls(path = "/PATH", glob = "*county*csv")
df <- read_csv(files, id = "path")

## or in one pipeline
df <- fs::dir_ls(path = "/PATH", glob = "*pattern*csv")%>%
  read_csv(id = "path")
```

## Challenge 12: Combining files with `readr` 

In the data folder, there is a sub-directory called penguins. This directory splits up the data from the `palmerpenguins` package into three data frames split by species. Turn these three files into one data frame. 

```{r}


```