---
title: "Part 1 Solutions"
author: "D-Lab"
output: html_document
---

```{r}
library(fs)
library(here)
library(tidyverse)
library(tibble)
```


## Challenge 1 

Here is a dataset inspired by an Economist/YouGov Poll about the importance of unemployment in the United States. 

```{r}
challenge_1 <- tibble(
  issue_importance = c("Very Important", "Somewhat Important",
                       "Not very Important", "Unimportant"),
  `18-29` = c(.59, .28, .08, .05),
  `30-44` = c(.66, .27, .06, .02),
  `45-64` = c(.69, .28, .03, .01),
  `65+` = c(.7, .27, .02, .01)
)

head(challenge_1)
```

Answer the following: 

1. How many observations are in the data set?

*16. While they are only 4 rows in this table, there are four categories of issue importance and four age groups. 4x4 = 16 total observations.*

2. How many columns are in the data set?

```{r}
ncol(challenge_1)
```

3. Is every column a variable?

*No! This data is messy. The first column is a variable that contains the universe of issue importance answers. However, the next five columns are all instances of the variable age.*


## Challenge 2

Create a new data frame object called `cats_and_dogs` that filters the `animals` data frame for all observations for which the animal type is a cat or a dog. Use the pipe operator to chain the functions together. 

```{r}
cats_and_dogs <- animals |> 
  filter(Type == "Cat"|Type == "Dog")
```
  
## Challenge 3

Let's put the verbs we've learned so far into action all at once. Answer the question "What are the total number of dog rescues for each borough?" by making a data frame called `borough_dog_rescues` and arranging the rows by borough in alphabetical order. 

```{r}
## Your Code here 
borough_dog_rescues <- animals |> 
  group_by(Borough) |>
  filter(Type == "Dog") |>
  count(Type) |>
  arrange(Borough)

```

## Challenge 4

Create a data frame that considers the average rescue cost, median rescue cost, and standard deviation for bird rescues for each property category year. Save the result of your workflow into a data frame object called `bird_stats`


```{r, include=F}
#YOUR CODE HERE
bird_stats <- animals |> 
  group_by(PropertyCategory, Year) |>
  filter(Type == "Bird") |>
  summarise(median_cost = median(RescueCost, na.rm = T),
            avg_cost = mean(RescueCost, na.rm = T), 
            sd_cost = sd(RescueCost, na.rm = T),
            .groups = "drop_last")

```


## Challenge 5

Add a column to the gapminder dataset that contains the continent's total population of each observation in a given year. For example, if the first observation was Afghanistan in 1952, the new column would contain the population of Asia in 1952. 

```{r}
gapminder_c6 <- gapminder |> 
  group_by(continent, year) |> 
  mutate(continent_pop = sum(pop))
``` 

  
## Challenge 6 

Use dplyr to: (a) add a column called `gdpPercap_diff` that contains the difference between the observation's `gdpPercap` and the mean `gdpPercap` of the continent in that year, (b) arrange the data frame by the column you just created, in descending order (so that the relatively richest country/years are listed first)

```{r}
gapminder_gdpPercap_diff <- gapminder_c6 |>  
  group_by(continent, year) |> 
   mutate(mean_continent_gdp = mean(gdpPercap),
          gdpPercap_diff = gdpPercap - mean_continent_gdp) |> 
  arrange(desc(gdpPercap_diff))
```