---
title: "Part 2 Solutions"
author: "D-Lab"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(fs)
library(here)
library(tidyverse)
library(tibble)
```

## Challenge 7 

The IGS poll also contains a question broken down by region (`cal_region.csv`). Build a pipeline that reads and tidies the data. For extra practice, make a similar graph to display the results. 

```{r}
## Full pipe here 
read_csv(here("data/cal_region.csv")) |> 
  pivot_longer(-leaving, 
               names_to = "region", 
               values_to = "percent_agreement") |> 
  ggplot(aes(x = leaving, y = percent_agreement, fill=region))+
  geom_col(position = 'dodge')+
  xlab("Moving Consideration Level")+
  ylab("Percent Agreement")+
  scale_fill_viridis_d()+
  ggtitle("Californians Consideration to Leaving the State by Region")

```

## Challenge 8

Edit the workflow presented above to split on the fifth character. Use `mutate` to replace the incorrect values with corrected and capitalized names.  

```{r}
## Here's a solution that takes advantage of stringr capabilities
## There are multiple ways to solve this problem
birds |> 
  pivot_longer(!c(county,year), 
               names_to = "sight",
               values_to = "n",
               values_drop_na = TRUE) |>
  separate(sight, c("type", "location"),sep = 5) |>
  mutate(type = str_replace_all(type, "warbl", "warbler")|>str_to_title(),
         location = if_else(str_detect(location, "Ne"), "Nest", "Air"))

### Here's a way using base string function 
birds |> 
  pivot_longer(!c(county,year), 
               names_to = "sight",
               values_to = "n",
               values_drop_na = TRUE) |>
  separate(sight, c("type", "location"),sep = 5) |>
  mutate(type = if_else(grepl("w", type, ignore.case = TRUE), "Warbler", "Eagle"),
         location = if_else(grepl("Ne", location), "Nest", "Air"))

### Here's another brute force way 
birds |>
  pivot_longer(!c(county,year), 
               names_to = "sight",
               values_to = "n",
               values_drop_na = TRUE) |>
  separate(sight, c("type", "location"),sep = 5) |>
  mutate(type = if_else(str_starts(type, "w"), "Warbler", "Eagle"),
         location = case_when(
           str_detect(location, "ir")~"Air",
           str_detect(location, "Ne")~"Nest",
           TRUE ~ location
         ))
```


## Challenge 9: Bulk Recoding Values 

Suppose you have a survey 

```{r}
set.seed(1234)
survey_raw_data <-tibble(
  ID = LETTERS[1:15],
  var1 = sample(1:15, replace = F),
  var2 = sample(16:30, replace = F),
  var3 = sample(31:45, replace = F)
)
glimpse(survey_raw_data)
```

Survey data is likely to contain errors. Perhaps the enumerators filled in a question incorrectly. Maybe you have a special value for missing data. As a subject matter expert, you realize that there are several fixes that need to be made globally. You put them into a data frame. 

```{r}
fixes <- tibble(
  ID = c("O", "G", "D", "D"),
  original_var = c("var1", "var2", "var2", "var3"),
  original_response = as.integer(c(1, 28, 19, 34)),
  correct_response = as.integer(c(1000, 2800, 1900, 3400))
)
glimpse(fixes)
```

Here we have a problem of updating several variables based on a set of potentially different conditions. Each variable needs to be updated based on the ID, but we do not need to update every variable/ID combination. 

Build a workflow that converts the incorrect answers to correct answers. A solution to this problem makes use of `left_join()`, `mutate()`, `select()`, and both `pivot_longer()` and `pivot_wider()`

```{r}
corrected_data <- survey_raw_data |> 
  pivot_longer(cols = -ID,
               names_to = "var_name",
               values_to = "original_value") |>
  left_join(fixes, by = c("ID", "var_name" = "original_var")) |>
  mutate(corrected_value = if_else(is.na(original_response),
                                   original_value,
                                   correct_response)) |>
  select(-original_value, -correct_response, -original_response) |>
  pivot_wider(names_from = "var_name",
              values_from = "corrected_value")
  

```

## Challenge 10: Combining files with readr 

In the data folder, there is a sub-directory called penguins. This directory splits up the data from the `palmerpenguins` package into three data frames split by species. Turn these three files into one data frame. 

```{r}
## Get files
files <- fs::dir_ls(path = here("data/penguins"), glob = "*species_*.csv")

## Combine data frames 
penguins_df <- files |> 
  read_csv(id = "path", show_col_types = FALSE)
```